Methods for Analyzing Models of Rod-Shaped Bacteria

### 2025-07-26 paper/main.tex
fix appendix
add ugly intermediate introduction

Bacteria exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.
The last decades have fundamentally challenged and changed our understanding of how bacteria retain
their shape.
Rod-shaped bacteria can grow by extending their circular part, or by inserting new material at the
tip of the rod.
Species such as E.coli and B.subtilis~\cite{Errington2020} fall under the first category while
S.pombe represents the latter.
New material is inserted in small bursts on the nanoscale in forms of patches, bands or
hoops~\cite{DePedro2003}.
Despite their differences in wall-size, \textit{B.subtilis} and \textit{E.coli} follow growth rules
which are comparable with respect to the extension of the rod~\cite{Chang2014}.

\paragraph{Experimental studies of mechanical properties}
\textbf{Study of Individual Bacteria}
\begin{itemize}
    \item \cite{Robert2014} growth of individual rods, distribution of parameters
    \item \cite{Koutsoumanis2013} inheritance of growth rates, growth parameter distributions
    \item \cite{Ursell2014} MreB acts againts curvature during growth of cell
    \item \cite{Takeuchi2005} cells in microchambers leads to plastic deformations
    \item \cite{Amir2014} bending stiffness, plastic + elastic deformations
\end{itemize}
\textbf{Collective Effects \& Self-Organization}
\begin{itemize}
    \item \cite{Trejo2013} mechanical response of bacterial colony
    \item \cite{Duvernoy2018} bacterial colonies form multilayers, interactions are polar
    \item \cite{vanGestel2015} van Gogh bundles form from interplay of 2 cell-types
\end{itemize}

@@

%###################################################################################################
\section*{Supplementary Material}

\renewcommand{\thesection}{S\arabic{section}}
\setcounter{section}{0}

>>>>

The last decades have fundamentally challenged and changed our understanding of how bacteria retain
their shape.
They exist in many physical shapes such as spheroidal, rod-shaped or spiral
\cite{Zapun2008,Young2006}.
Rod-shaped bacteria can grow by extending their circular part, or by inserting new material at the
tip of the rod.
Species such as \textit{E.coli} and \textit{B.subtilis}~\cite{Errington2020} fall under the first
category while
S.pombe represents the latter.
New material is inserted in small bursts on the nanoscale in forms of patches, bands or
hoops~\cite{DePedro2003}.
Despite their differences such as wall-size or length \textit{B.subtilis} and \textit{E.coli} follow
growth rules which are comparable with respect to the extension of the rod~\cite{Chang2014}.

\paragraph{Experimental studies of mechanical properties}
It was long believed that bacteria lack cytoskeletal filaments but with the end of the last
century, it was discovered that the MreB protein fills this role since it "[..] forms an actin-like
cytoskeleton in bacteria [..]" \cite{Erickson2001}.
It can polymerize and form filaments which behave similar to actin microfilaments \cite{Dersch2020}.
Studies of its crystal structure further showed the similarity to actin \cite{vandenEnt2001} and the
MreB, MreC and MreD proteins were identified as homologues of actin \cite{Lowe2017_lj}.\
Jones et al. studied the species of different bacterial sub-kingdoms which have the MreB
gene~\cite{Jones2001}.
They concluded: "Many of the organisms are rod shaped, similar to \textit{B. subtilis} and
\textit{E. coli}.
Organisms with more complex shapes, including curved, filamentous, and helical bacteria, were all
abundantly represented." \cite{Jones2001}.
Furthermore, \cite{Wachi1987} found that mutants of \textit{E.Coli} which create defective MreB
proteins will be spherical instead of their natural rod-shaped form.

% \begin{itemize}
%     \item \cite{Robert2014} growth of individual rods, distribution of parameters
%     \item \cite{Koutsoumanis2013} inheritance of growth rates, growth parameter distributions
%     \item \cite{Ursell2014} MreB acts againts curvature during growth of cell
%     \item \cite{Takeuchi2005} cells in microchambers leads to plastic deformations
%     \item \cite{Amir2014} bending stiffness, plastic + elastic deformations
% \end{itemize}
Other experiments found that individual rods of \textit{E.Coli} grow exponentially until the
division event which is triggered by a size-sensing rather than a time-sensing
mechanism~\cite{Robert2014}.
These results also showed that the growth parameters were randomly distributed but did not quantify
said distribution.
Using $220$ \textit{Salmonella enterica}, Koutsoumanis et al. analysed the distribution of division
times and was able to quantify their distributions~\cite{Koutsoumanis2013}.
They found that while the initial division times were longer, there was no correlation between
division times of mother and daughter cells.
The first finding can be explained by the bacterias lag-phase~\cite{Bertrand2019} which is a
temporary period during which bacteria do not grow.
The second observation is an example of principle of thought which underlyies guides this work:
Cells are individual agents with individual parameters and thus individual behaviour and need to be
treated as such.
This is especially important if we want to determine parameters of our model for individual agents.

% \paragraph{Collective Effects \& Self-Organization}
% are emergent phenomena that result from the combined interactions of multiple cells.
% Although they do not allow us to fundamentally understand, the individual behaviour of cells
% directly, they can provide a way to falsify our results and thus further guide our modeling
% approach.
% \cite{Trejo2013} investigated the mechanical response of a tightly packed colony of \textit{Bacillus
% Subtilis}.
% \cite{Trejo2013} studied the formation of macroscopic wrinkles when growing bacteria in a confined
% space.
% They were able to model the strain and stress of the collection of bacteria which requires that
% cells are connected by a (to first order) spring-like force.
% The developing wrinkles could then be described by a buckling instability.
% 
% Duvernoy et al. used laser techniques and force microscopy to investigate adhesion of individual
% \textit{E.coli} and \textit{Pseudomonas aeruginosa} cells~\cite{Duvernoy2018}.
% They found that the adhesion forces of the rods are polar, resulting in mechanical tension which in
% turn determines daughter cell arrangement.
% They measured the size of the grown bacterial colonies, fitted ellipses to their shapes and found
% that adhesion and polar adhesion in particular resulted in a greater difference between the long and
% short axis of the fitted ellipse and thus a more oval shape rather than circular (WT).
% By extending the work of \cite{Grant2014}, they showed, that the critical value of transitioning
% from a monolayer to multiple layers of bacteria, depends on the adhesive strength and polarity of
% the interaction.
% \cite{Grant2014} described the growing bacterial colony as a disc with radius $R$ which is pressed
% into the agarose slab.
% The authors created a purposely-designed simulation in `C++` in order to model the effects of the
% gel on the collection of bacteria.
% 
% Van Gestel et al. explored how \textit{B.subtilis} migrates over a surface by forming multicellular
% structures~\cite{vanGestel2015}.
% The cells organize themselves into "van Gogh bundles" of cells which are tightly aligned in chains
% and form filamentous loops.
% This phenomenon occurs at the border of the bacterial colony and the migration is driven by two
% phenotypically different cell types~\cite{Lpez2010}.
% \begin{itemize}
%     \item \cite{Trejo2013} mechanical response of bacterial colony
%     \item \cite{Duvernoy2018} bacterial colonies form multilayers, interactions are polar
%     \item \cite{vanGestel2015} van Gogh bundles form from interplay of 2 cell-types
% \end{itemize}

@@

\appendix
%###################################################################################################
\section*{Supplementary Material}

\renewcommand{\thesection}{S\arabic{section}}
